---
title: "p8105_hw2_yl5828"
author: "Leah Li yl5828"
date: "2025-10-01"
output: github_document
editor_options: 
  chunk_output_type: console
---


```{r message = FALSE}
# Load dataset
library(tidyverse)
library(dplyr)
library(readr)
library(lubridate)
```


## Problem 1

### Clean file pols-month

```{r read and clean pols_month}
pols_month_df =
  read_csv("./fivethirtyeight_datasets/pols-month.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names() %>%
  separate(mon, into = c("year", "month", "day"), sep = "-") %>%
  mutate(
    month=recode(month, `01` ="January", `02`= "February", `03`= "March", `04`= "April", `05`= "May",`06`= "June",`07`= "July",`08`= "August",`09`= "September",`10`= "October",`11`= "November",`12`= "December"),
    president = if_else(prez_gop == 1, "gop", "dem"),
    year = as.numeric(year)
         )%>%
  select(-prez_dem, -prez_gop, -day)

pols_month_df
```

### Clean file snp

```{r read and clean snp}
snp_df =
  read_csv("./fivethirtyeight_datasets/snp.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names() %>%
  separate(date, into = c("month", "day", "year"), sep = "/") %>%
  mutate(
    month=recode(month, `1` ="January", `2`= "February", `3`= "March", `4`= "April", `5`= "May",`6`= "June",`7`= "July",`8`= "August",`9`= "September",`10`= "October",`11`= "November",`12`= "December"),
    year = if_else(year > 25, 1900 + as.numeric(year), 2000 + as.numeric(year)),
    month = factor(month, levels = month.name, ordered = TRUE))%>%
  select(year, month, close) %>%
  arrange(year,month)

snp_df
```

### Clean file unemployment

```{r read and clean unemployment}
unemployment_df = 
  read_csv("./fivethirtyeight_datasets/unemployment.csv", na = c("NA", ".", "")) %>%
  janitor::clean_names() %>%
  pivot_longer(
    jan:dec,
    names_to = "month", 
    values_to = "unemployment") %>%
  mutate(
    month=recode(month, `jan` ="January", `feb`= "February", `mar`= "March", `apr`= "April", `may`= "May",`jun`= "June",`jul`= "July",`aug`= "August",`sep`= "September",`oct`= "October",`nov`= "November",`dec`= "December" ))

unemployment_df
```

### Join datasets

```{r Join datasets}
q1join_df = 
  inner_join(pols_month_df, snp_df, by = c("year", "month")) %>%
  inner_join(unemployment_df, by = c("year", "month"))

q1join_df
```

### Dataset Explanation

The dataset `pols_month_df` include variables as follows: `r paste(names(pols_month_df), collapse = ", ")`. And the dataset has **`r nrow(pols_month_df) `** rows and **`r ncol (pols_month_df)`** columns. 

The dataset `snp_df` include variables as follows: `r paste(names(snp_df), collapse = ", ")`. And the dataset has **`r nrow(snp_df) `** rows and **`r ncol (snp_df)`** columns. 

The dataset `unemployment_df` include variables as follows: `r paste(names(unemployment_df), collapse = ", ")`. And the dataset has **`r nrow(unemployment_df) `** rows and **`r ncol (unemployment_df)`** columns. 

The resulting dataset `q1join_df` include variables as follows: `r paste(names(q1join_df), collapse = ", ")`. And the dataset has **`r nrow(q1join_df) `** rows and **`r ncol (q1join_df)`** columns. The range of years is (**`r range(q1join_df$year)`**).



## Problem 2

### Clean dataset Mr.Trash Wheel

```{r read and clean Mr.Trash Wheel}
mr.trash_df = 
  readxl::read_excel("./Mr.Trash Wheel/202509 Trash Wheel Collection Data.xlsx", sheet="Mr. Trash Wheel", range="A2:M709", na = c("NA", ".", ""))%>%
  janitor::clean_names() %>%
  drop_na(dumpster) %>%
  mutate(
    sports_balls = as.integer(round(sports_balls)),
    year = as.numeric(year),
    wheel_name = "Mr.Trash"
  )

mr.trash_df
```

### Clean dataset Professor Trash Wheel

```{r read and clean Professor Trash Wheel}
professor_trash_df = 
  readxl::read_excel("./Mr.Trash Wheel/202509 Trash Wheel Collection Data.xlsx", sheet="Professor Trash Wheel", range="A2:M134", na = c("NA", ".", ""))%>%
  janitor::clean_names() %>%
  drop_na(dumpster)%>%
  mutate(wheel_name = "Professor")

professor_trash_df
```

### Clean dataset Gwynnda Trash Wheel

```{r read and clean Gwynnda Trash Wheel}
gwynnda_trash_df = 
  readxl::read_excel("./Mr.Trash Wheel/202509 Trash Wheel Collection Data.xlsx", sheet="Gwynns Falls Trash Wheel", range="A2:L351", na = c("NA", ".", ""))%>%
  janitor::clean_names() %>%
  drop_na(dumpster)%>%
  mutate(wheel_name = "Gwynnda")

gwynnda_trash_df
```

### Join the three wheels together

```{r}
all_wheels_combined <- bind_rows(mr.trash_df, professor_trash_df, gwynnda_trash_df)

all_wheels_combined
```

### Dataset Explanation

```{r Professor Trash Wheel Total Weight}
prof_weight <- all_wheels_combined %>%
  filter(wheel_name == "Professor") %>%
  summarise(total_weight = sum(weight_tons, na.rm = TRUE)) %>%
  pull(total_weight)

print(prof_weight)
```

The total weight of trash collected by Professor Trash Wheel was **`r round(prof_weight, 2)`** tons.

```{r Gwynnda Trash Wheel Cigarette Butts}
gwynnda_butts <- all_wheels_combined %>%
  filter(wheel_name == "Gwynnda", month == "June", year == 2022) %>%
  summarise(total_butts = sum(cigarette_butts, na.rm = TRUE)) %>%
  pull(total_butts)

print(gwynnda_butts)
```

The total number of cigarette butts collected by Gwynnda in June of 2022 was **`r format(gwynnda_butts, big.mark = ",")`**. 

The dataset contains **`r nrow(all_wheels_combined)`** observations of trash collection events for `r n_distinct(all_wheels_combined$wheel_name)` different trash wheels, spanning from `r min(all_wheels_combined$year)` to `r max(all_wheels_combined$year)`. Each observation records the contents of a single dumpster, identified by the name of the trash wheel (e.g., 'Mr. Trash Wheel', 'Gwynnda'). The dataset `mr.trash_df` include **`r nrow(mr.trash_df) `** dumpsters. The dataset `professor_trash_df` include **`r nrow(professor_trash_df) `** dumpsters. The dataset `gwynnda_trash_df` include **`r nrow(gwynnda_trash_df) `** dumpsters. The combined dataset has the following variables: **`r paste(names(all_wheels_combined), collapse = ", ")`**. Key variables include the total weight_tons for each haul, as well as counts of specific items like plastic_bottles, polystyrene, and cigarette_butts.


## Problem 3

### Read in dataset Zip_codes and ZORI

```{r Read in zip_codes and zori}
zip_codes <- read_csv("./zillow_data/Zip Codes.csv")
zori <- read_csv("./zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv")
```

### Clean dataset Zip_codes

```{r Clean zip_codes}
zip_codes_clean <- zip_codes %>%
  janitor::clean_names() %>%
  select(zip_code, county, neighborhood) %>%
  distinct()
```

### Clean dataset ZORI

```{r Clean ZORI}
zori_clean <- zori %>%
  rename(zip = RegionName)%>%
  pivot_longer(
    cols = !RegionID:CountyName,  
    names_to = "date",
    values_to = "zori"
  ) %>%
  mutate(
    date = ymd(date),
    zip = as.integer(zip)) %>%
  select(zip, CountyName, date, zori)
```

### Merge Zip_codes with ZORI.

```{r Merge zori_clean and zip_codes_clean}
zori_merged <- zori_clean |>
  left_join(zip_codes_clean, by = c("zip" = "zip_code"),relationship = "many-to-many") %>%
  select(date,zip,CountyName,neighborhood,zori)
```

### Briefly describe the resulting zori_merged dataset.

The resulting _tidy_ dataset **`zori_merged`** contains **`r nrow(zori_merged)`** total observations. The dataset includes **`r n_distinct(zori_merged$zip)`** unique ZIP codes and **`r n_distinct(na.omit(zori_merged$neighborhood))`** unique neighborhoods across the five boroughs of New York City.

There are some ZIP codes apeear in the ZIP code dataset but not in the Zillow Rental Price dataset, such as,  **`r paste(head(setdiff(zip_codes_clean$zip_code, zori_clean$zip), 5), collapse = ", ")`**.  

These ZIP codes are often excluded because they may correspond to invalid information. For example, the ZIP codes are in fact **non-residential areas** (large commercial or industrial zones) or **P.O. boxes** numbers. 



### Rental Price Fluctuation

```{r Generate top 10 drop during COVID}
covid <- zori_merged %>%
  filter(date %in% c("2020-01-31", "2021-01-31")) %>%
  pivot_wider(
    names_from = date,
    values_from = zori,
    values_fn = mean) %>%
  mutate(drop = `2021-01-31` - `2020-01-31`) %>%
  arrange(drop) %>%
  slice_head(n = 10) %>%
  select(zip, CountyName, neighborhood, `2020-01-31`, `2021-01-31`, drop)

covid
```

### Comments on Rental Price Fluctuation

From the table, we could observe that rental prices fluctuated dramatically during the COVID-19 pandemic.  
Comparing January 2021 to January 2020 across all ZIP codes, the largest decline was in ZIP 10007 Lower Manhattan from $6334 to $5422 resulting a drop of $913. Second large drop happened in ZIP 10069 with about $748, following with a $714 drop in ZIP 10009 Lower East Side. 

Other borough neighborhoods, such as Queens, Bronx, Staten Island, are not in the top 10, suggesting less fluctuations in these areas.
